---
title: "Changes in Thresholds for<br>Hospital Admission Via Emergency Department"
subtitle: "A National-level analysis<br><br><br>DRAFT REPORT<br>Winter 2025"
author:
  - name: Steven Wyatt
    email: swyatt@nhs.net
    corresponding: TRUE
    affiliation: The Strategy Unit
  - name: Andrew Jones
    email: swyatt@nhs.net
    corresponding: FALSE
# date: 2024-11-29
format:
  html:
    html-table-processing: none
    # code-fold: true
    # code-tools: true
    embed-resources: true
    grid:
      margin-width: 240px
  pdf: default
reference-location: margin
citation-location: margin
toc: true
toc-title: Contents
toc-location: left
# theme: journal
fig-cap-location: top
number-sections: false
# backgroundcolor: '#FFE0B5'
title-block-banner: '#f8be06'
title-block-banner-color: '#242322'
# mainfont: Gill Sans MT
# mainfont: Helvetica
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

library("gt")
library("here") 
library("arrow")
library("dplyr")
library("purrr")
library("readr") 
library("tidyr")
library("tibble")
library("tsibble")
library("forcats")
library("ggplot2") 
library("janitor")
library("stringr")
library("gtExtras") 
library("lubridate")


plot_condition_over_time <- function(df) {
  df |>
    ggplot(aes(as_date(year_quarter), n, group = diag, col = diag)) +
    geom_line() +
    geom_point(size = 0.6) +
    geom_blank(aes(y = 0)) +
    facet_wrap(vars(diag), scales = "free_y") +
    theme_minimal() +
    theme(
      strip.text = element_text(size = 7),
      legend.position = "none",
      axis.text = element_text(size = 7),
      axis.title = element_text(size = 7),
    )
}

```

# Introduction

# Methods

## Data source

This analysis relied on data from the Emergency Care Data Set (ECDS). Each ECDS record represents a patient attendance at an urgent or emergency care department. We selected the subset of records coded as unplanned attendances at Type 1 emergency departments (EDs), with an arrival date between 1st April 2019 and 31st March 2025 (inclusive).

We removed records where the patient left the ED before treatment was provided; where the patient died during an attendance; and where the individual was found to be dead on arrival at the ED.

## Data quality and use of sampling

Whilst the ECDS is recognised as a high-quality data source, there is yet considerable variation in data quality, both across provider sites and over time. If overlooked, this variation could adversely affect the validity of our results.

We were particularly concerned, here, about the completeness and consistency of variables that indicate the clinical features and severity of a patient's condition. These fields - *diagnosis*, *chief complaint*, and *acuity* - are likely to be the strongest predictors of admission and therefore have the greatest influence on our results. If we were to use a dataset with sizeable inconsistency in the coding of these fields (non-constant risks) over time, this could distort - and invalidate - our results. To increase the chances that our findings would be valid, we conducted the analysis on records from a sample of sites whose data met our quality requirements.[^1]

[^1]: This sample was therefore non-random.

Whilst sampling should improve the validity of our results, we can only make *general statements* about changes in admission thresholds over time if our sample of sites - and the patients attending these - is *representative* of EDs nationwide. In other words, on aggregate, the sites and patients covered by our sample should have similar characteristics to those across England.

There is a trade-off to be made, however. The larger our sample of sites, the more likely it is to be representative of the population of sites, and the more likely it will be that our findings are generalisable. However, the larger we make our sample, the more we compromise on data quality, which will likely decrease the validity of our results.

Details of our approach to this trade-off, as well as our data quality criteria, sampling methods, and the measures of representation, can be found below.

#### Sample criteria

We identified 138 sites that submitted data to ECDS throughout the study period (2019/20 to 2024/25) and that had a consistent site code. These 138 EDs will represent the total population of providers that we might have used in this analysis had data quality been universally high.

As explained above, we viewed the completeness and consistency across *diagnosis*, *acuity*, and *chief complaint* fields as being particularly important to this study. From the population of 138 sites, only 20 had completeness rates for these variables that exceeded 75% in every quarter of the period studied.

To ensure yet greater consistency, we excluded a further 3 sites (of the selection of 20) where the completion rate of diagnosis decreased, on average, by more than 3% per year. And we excluded one further site that showed two sizeable step changes in the completion of the diagnosis field. This left us with a working sample of 16 sites.

<br>

```{r}
#| label: fig-diag-completion
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "<br><br>"
#| column: page
#| fig-width: 12
#| fig-asp: 0.618
#| out-width: "85%"


readRDS(here("data", "251204_qdf_diag_consistency.rds")) |>
  tidyr::complete(der_provider_site_code, year_quarter) |> 
  mutate(p = if_else(is.na(p), 0, p)) |> 
  ggplot(aes(as_date(year_quarter), p, group = diag_descr_snomed, col = diag_descr_snomed))+
  geom_line(alpha = .5, col = "dodgerblue2")+
  geom_smooth(method = "lm", se = F, alpha = .3)+
  geom_point(size = 0.6, col = "dodgerblue2")+
  geom_blank(aes(y = 0))+
  facet_wrap(vars(der_provider_site_code))+
  # facet_wrap(vars(der_provider_site_code), scales = "free_y")+
  theme_minimal()+
  scale_y_continuous(labels = scales::percent)+
  theme(
    strip.text = element_text(size = 8),
    legend.position = "none",
    axis.text = element_text(size = 6),
    axis.title = element_text(size = 8),
  )+
  labs(
    x = "\nDate (quarter commencing)",
    y = "Percentage of missing diagnosis values\n"
    )


```

#### Continuity of outcome variable

Finally, it was necessary to verify that our intended outcome variable (whether the patient was admitted or not) was complete throughout the study period, for each site in our working sample. At the same time, we established that any change in the crude rate of admission, by site, did not exceed +/-10% from any one quarter to the next. This might have been indicative of a data quality issue.

```{r}
#| label: fig-admitted-completion
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "<br><br>"
#| column: page
#| fig-width: 12
#| fig-asp: 0.618
#| out-width: "85%"

readRDS(here("data", "251204_dfq_outcome_ts.rds")) |> 
  ggplot(aes(as_date(year_quarter), p_adm))+
  geom_line()+
  geom_point(size=.6)+
  geom_blank(aes(y= 0))+
  theme_minimal()+
  theme(
    strip.text = element_text(size = 6),
    axis.text = element_text(size = 6),
    axis.title = element_text(size = 9)
    )+
  scale_y_continuous(labels = scales::percent, breaks = c(0, 0.2, 0.4))+
  # facet_wrap(vars(str_c(der_provider_site_code, "\n", site_name)))+
  facet_wrap(vars(der_provider_site_code))+
  labs(x = "\nDate (quarter commencing)", y = "% patients admitted\n")




```

### Sites in the sample

The 16 EDs meeting the data quality criteria above are named in table X. Data from these sites, which captures more than 7 million attendances over 6 years, will be used in the following analysis. Table y shows how the characteristics of EDs in the sample compare to the characteristics of EDs nationally.

The sample of EDs is slightly skewed towards smaller sites, and there is no representation of sites in London or North West England. However, the casemix of patients attending the EDs in our sample was broadly representative of the casemix of all patients who attended EDs over the period.

When we relaxed the completeness threshold for our key variables from 75% to 70%, we saw no improvement in representation. If we were to continue to lower the threshold, in search of greater representation, we may affect the validity of our results.

::: panel-tabset
## Sites in sample

```{r}
#| label: tbl-sample-sites
#| tbl-cap: "A list of sites used in this analysis."
#| echo: false
#| message: false
#| column: page


readRDS(here("data", "251204_df_sample_sites.rds")) |>
   gt() |>
  fmt_auto() |>
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels(c(site_name, trust_name))
  ) |>
  tab_options(
    data_row.padding = px(1),
    table.font.size = "small"
  )



```

## How representative is this sample?

```{r}
#| label: tbl-representation
#| tbl-cap: "Table comparing selected characteristics of the sample of sites used in this analysis with the characteristics of all qualifying sites across England."
#| echo: false
#| message: false
#| column: page

readRDS(here("data", "251204_table_representation.rds")) |> 
    gt() |> 
  opt_row_striping(row_striping = F) |> 
  fmt_auto() |> 
  cols_label(
    rep_diff = "Representation:<br>Sample vs Pop",
    .fn = md
  ) |> 
  tab_style(
    style = list(
      cell_text(transform = "capitalize", weight = "bold")
    ),
    # different location
    # locations = cells_column_labels(everything())
    locations = cells_column_labels(1:3)
  ) |> 
  fmt_percent(rows = c(2:24), decimals = 0) |>
  fmt_percent(columns = 4, decimals = 0) |>
  # sub_values(
  #   rows = 1,  fn = function(x) x >= 0 & x < 1,
  #            replacement = "") |> 
  tab_row_group(
    label = "Patient demographics:",
    rows =  str_detect(variable, "depriv|region_")
  ) |> 
  tab_row_group(
    label = "Case-mix-related:",
    rows =  str_detect(variable, "age|arrive_|admitted")
  ) |> 
  tab_row_group(
    label = "Provider-related (23/24):",
    rows =  str_detect(variable, "att|also|4h|12h")
  ) |> 
  tab_row_group(
    label = "",
    rows =  str_detect(variable, "n_sites|23")
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups(groups = c(1, 2, 3, 4))
  ) |> 
  gt_highlight_cols(sample, fill = "grey", alpha = 0.1) |> 
  tab_options(
    data_row.padding = px(1),
    table.font.size = "small"
  )

  

```
:::

## Data processing

#### **Addressing incompleteness**

On average, XX% of records in our sample had missing values in *at least one* of the fields we had selected for our model. Since our intended logistic regression model would not accept missing values, we had the choice to either: 1) exclude records with missing values; 2) impute missing values; or 3) assign missing values to an explicit missing category.[^2]

[^2]: This was only possible as the variables with missing values were exclusively categorical.

We opted for the third approach, which allowed us to make use all the records in our study dataset without the need to employ an additional (and potentially complex) imputation algorithm. This approach also allows the model to make use of the missingness signals. Importantly here, we haven’t made the assumption that data were missing at random. We believed, instead, that the missingness was more likely to be due to failures of systems and processes.[^3]

[^3]: If, however, the missingness was random, there is a risk that models we build could incorporate a false signal. This signal would compete with other variables for explanatory power, which may ultimately distort the coefficients of interest.

#### Creating new variables

Further to the treatment of missing values, we chose to synthesise a variable that captured the clinical features of a patient's condition. The reasons for this were largely related to the fact that the *primary diagnosis* variable, unmodified, consisted of a large number of categories (9,800+), which would translate into a similarly large number of predictor variables. If we could substantially reduce the number of predictors whilst minimising the loss of information, we would be more likely to end up with a model that was more capable of distinguishing between genuine effect and random variation.

To produce this new variable we used information from the diagnosis and chief complaint variables, together with the 3-tier hierarchy from the “ECDS Refset MAX specification”, created by NHSE.

We first created a frequency table of primary diagnoses in our sample, and noted the 30 most frequent SNOMED codes for use in our new variable. To this list, we added two SNOMED diagnosis codes related to acute COVID-19 infection.[^4]

[^4]: These two codes are mapped to one condition (COVID-19).

Having removed activity with these diagnosis codes from our sample, we then identified the 20 most frequent chief complaint codes and noted these.

We mapped the remaining chief complaint codes to the ECDS Refset diagnosis table, and grouped conditions at the highest (most general) level. Whilst there were 36 instances where the SNOMED code was labelled as “deprecated” when mapped to the Refset, we were able to categorize these manually with the aid of archived material.

We therefore captured the condition (where given) of all patients in the sample, at various levels of detail, in 66 categories. On examining the trends for these 66 categories, we saw evidence of codes being re-classified or dropped entirely. We attempted to reconcile these discontinuities, by:

-   Uniting different codes for urinary tract infections

-   Uniting different codes for injuries of the upper, and lower, extremities

-   Moving lower limb pain and upper limb pain from Level 2 to musculoskeletal conditions in Level 1.

-   Moving drug and alcohol abuse from Level 1 into substance misuse in Level 2.

These changes left us with 62 categories indicating the clinical features of a patient’s condition. The 62 categories, and their timeseries, can be seen in @fig-diag-ts-l3, @fig-diag-ts-l2, and @fig-diag-ts-l1. The proportion of attendances covered by each tier is shown in ....

::: panel-tabset
## Diagnosis tier 3 (most precise)

```{r}
#| label: fig-diag-ts-l3
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "<br><br>"
#| column: page
#| fig-width: 12
#| fig-asp: 0.618
#| out-width: "85%"

df_preplot_condition_over_time <- readRDS(here("data", "251211_dfq_condition_over_time.rds"))

df_preplot_condition_over_time$data[[1]] |> 
  plot_condition_over_time()

```

## Diagnosis tier 2

```{r}
#| label: fig-diag-ts-l2
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "<br><br>"
#| column: page
#| fig-width: 12
#| fig-asp: 0.618
#| out-width: "85%"


df_preplot_condition_over_time$data[[2]] |> 
  plot_condition_over_time()

```

## Diagnosis tier 1 (most general)

```{r}

#| label: fig-diag-ts-l1
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "<br><br>"
#| column: page
#| fig-width: 12
#| fig-asp: 0.618
#| out-width: "85%"


df_preplot_condition_over_time$data[[3]] |> 
  plot_condition_over_time()

```
:::

```{r}
#| label: tbl-diag-propos
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "<br><br>"
#| column: page
#| fig-width: 12
#| fig-asp: 0.618
#| out-width: "85%"

df_preplot_condition_over_time |> 
  select(level, p) |> 
  gt() |> 


```

Moving on to other variables, we created unique levels of the acuity variable – merging two or more of the original levels – for 3 EDs, since, for these sites, we observed erratic variation in the acuity levels over time.

We created a new *referral source* variable consisting of six levels, which combined categories from the *attendance source* field. Finally, we created 3 new binary variables indicating 1) whether the patient arrived at ED by ambulance (of any type); 2) whether the patient arrived at outside daytime hours (or 8pm - 8am); and 3) whether the patient arrived at a weekend.

::: panel-tabset
## Separate

```{r}
#| label: fig-acuity-ts-sep
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "<br><br>"
#| column: page
#| fig-width: 12
#| fig-asp: 0.618
#| out-width: "85%"


readRDS(here("data", "251204_dfq_acuity_over_time.rds")) |> 
  #
  mutate(acuity = as.character(acuity)) |> 
  mutate(acuity = case_when(
    site_name == "Great Western Hospital A&E" &
      acuity %in% c("Standard", "Urgent", "Non-urgent") ~ "NonUrg/Std/Urg",
    site_name == "Worthing Hospital" &
      acuity %in% c("Standard", "Non-urgent") ~ "NonUrg/Std",
    site_name == "Leicester Royal Infirmary" &
      acuity %in% c("Standard", "Urgent") ~ "Std/Urg",
    T ~ acuity
  )) |> 
  #
  count(site_name, year_quarter, acuity, wt = n) |> 
  group_by(site_name, year_quarter) |> 
  mutate(p = n/sum(n)) |> 
  ungroup() |> 
  ggplot(aes(year_quarter, p, col = acuity, group = acuity))+
  geom_line()+
  geom_point(size = .9)+
  geom_blank(aes(y=0))+
  #facet_wrap(vars(site_name), scales = "free_y")+
  facet_wrap(vars(site_name))+
  theme_minimal()+
  theme(
    legend.position = "top",
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    axis.text = element_text(size = 6),
    axis.title = element_blank(),
    strip.text = element_text(size = 8),
  )+
  guides(colour = guide_legend(nrow = 1))

```

## Merged

```{r}
#| label: fig-acuity-ts-merged
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "<br><br>"
#| column: page
#| fig-width: 12
#| fig-asp: 0.618
#| out-width: "85%"


readRDS(here("data", "251204_dfq_acuity_over_time.rds")) |> 
  count(site_name, year_quarter, acui3, wt = n) |> 
  group_by(site_name, year_quarter) |> 
  mutate(p = n/sum(n)) |> 
  ungroup() |> 
  ggplot(aes(year_quarter, p, col = acui3, group = acui3))+
  geom_line()+
  geom_point(size = .6)+
  geom_blank(aes(y=0))+
  facet_wrap(vars(site_name), scales = "free_y")+
  theme_minimal()+
  theme(
    legend.position = "top",
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    axis.text = element_text(size = 6),
    axis.title = element_blank(),
    strip.text = element_text(size = 8),
  )+
  guides(colour = guide_legend(nrow = 1))

```
:::

## Model specification

We fitted a logistic generalised additive model (logistic GAM) to the processed data, with our binary outcome variable being whether the patient was admitted or not. Our variable of interest was the quarter of the financial year that a patient attended. We also included variables to adjust for casemix (as far as possible), namely: *age* (as a spline-based smooth) interacting with *sex*; *acuity*; the new, grouped primary diagnosis variable; whether the patient arrived by ambulance or not; and the *referral source*. We incorporated variables to account for the ethnicity and socioeconomic status of patients.[^5]^,^ [^6] In addition, we included time-related variables to control for weekday or weekend attendances, and whether the attendance was during the night (8 pm to 8 am).

[^5]: The ethnicity variable contained 18 categories which broadly corresponded to the [high-level groups used by the ONS in the 2021 census](https://www.ethnicity-facts-figures.service.gov.uk/style-guide/ethnic-groups/).

[^6]: The Index of Multiple Deprivation quintile covering the area in which a patient lived was used a proxy for socioeconomic status.

The ED site was the final variable in the model. As graphical evidence indicated that different sites might interpret a patient's acuity level in different ways (and the acuity variable would likely be highly influential), we included an interaction between ED site and acuity.

We intentionally omitted contextual factors such as the level of crowding and the inpatient occupancy at the time a patient was treated. Whilst these variables are challenging to engineer for an extended time period, we also wanted to leave these factors as possible explanations for any changes seen in admission thresholds **(less relevant excuse for quarterly data??)**. If we were to adjust for such effects they would not (in theory) be visible in the time series.

We fitted the model to a random sub-sample of X million records.

## Reproducibility

The computer code used in this analysis can be found a [GitHub repository](https://github.com/The-Strategy-Unit/tracking_adm_thresholds) (permissions currently required).

# Results

The aim of this analysis was to describe national-level changes in threshold for hospital admission via ED. We therefore used a statistical model to estimate the casemix-adjusted odds of admission over time. The model coefficients corresponding to the each quarter of every financial year give us these odds, and we show these coefficients in Fig X.

Note confidence intervals - HOW TO INTERPRET which may change.

The results can dotted line represents odds are equal to those of Q1 of 2019/20. We might divide the time period into three parts: the pre-pandemic period; the pandemic period; and the post-pandemic period. The first point to note is that these estimates generally suggest that the odds of admission during the pandemic period were 10-15% higher than in the other periods.

The odds of admission pre- and post-pandemic are similar (to be interpreted with full model).

Values of the other (parametric) model coefficients, and the model perfomance statistics, are given in Appendix X

```{r}
#| label: fig-odds-quarter
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "<br><br>"
#| column: page
#| fig-width: 12
#| fig-asp: 0.618
#| out-width: "85%"

# readRDS(here("data", "251204_dfq_coeff_II.rds")) |>
# readRDS(here("data", "2512xx_dfq_coeff_II.rds")) |>
#readRDS(here("data", "251208_dfq_coeff_II.rds")) |>
#readRDS(here("data", "2512yy_dfq_coeff_II.rds")) |>
readRDS(here("data", "251211_dfq_coeff_II.rds")) |>
  filter(str_detect(term, "year")) |>
  mutate(term = str_remove(term, "year_quarter")) |>
  mutate(row = row_number()) |>
  ggplot() +
  geom_hline(yintercept = 1, lty = "dashed", colour = "grey20", alpha = 0.5) +
  geom_pointrange(aes(term, odds, ymin = lci, ymax = uci), col = "grey40", stroke = NA) +
  # geom_smooth(aes(row, odds), se = F)+
  theme_minimal() +
  # coord_flip() +
  scale_y_log10(
    limits = c(0.87, 1.24)
    # breaks = c(0.9, 1, 1.1),
    # labels = c(
    #   # "Half\nas likely in\n2023/24",
    #   # "Equally\nas likely in\n2023/24",
    #   # "2x\nas likely in\n2023/24",
    #   "90%\nas likely as\nav. occupancy",
    #   "Equally\nas likely as\nav. occupancy",
    #   "110%x\nas likely as\nav. occupancy"
    # )
  ) +
  scale_x_discrete(
    # breaks = seq(1, 23, 2)
    # labels = 
  )+
  theme(
    # axis.title.y = element_blank(),
    axis.text = element_text(size = 7),
    axis.title.y = element_text(size = 9),
    axis.title.x = element_blank(),
    # plot.margin = margin(5, 20, 5, 5),
    # panel.grid.minor = element_blank(),
  ) +
  labs(
    y = "Case-mix-adjusted odds of admission (log scale)\nReference category: 2025 Q4\n"
  ) 

```

For context: Diagnosis.

```{r}
#| label: fig-odds-diacuity
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "<br><br>"
#| column: page
#| fig-height: 15
#| fig-asp: 1.618
##| out-width: "100%"
# #| fig-width: 12


readRDS(here("data", "251204_dfq_coeff_II.rds")) |>
  filter(str_detect(term, "iag|acu")) |>
  mutate(grp = str_sub(term, 1, 3)) |> 
  # count(grp)
  # filter(str_detect(term, "diag")) |>
  # filter(str_detect(term, "walk"))
  # arrange(desc(odds)) |>
  # slice_sample(n = 30) |>
  ggplot() +
  geom_hline(yintercept = 1, lty = "dashed", colour = "grey20", alpha = 0.5) +
  geom_pointrange(aes(reorder(term, odds), odds, ymin = lci, ymax = uci, col = grp),  stroke = NA) +
  theme_minimal() +
  coord_flip() +
  scale_y_log10() +
  theme(
     axis.title.x = element_text(size = 9),
    axis.title.y = element_blank(),
    legend.position = "none"
  ) +
  labs(
    y = "\nCase-mix-adjusted odds of admission (log scale)"
  )+
  facet_wrap(vars(grp), scales = "free_y", ncol = 1)


```

# Discussion

Text here

# Appendix

#### Performance measures

The fitted model scored an AUROC of 0.866 and a Brier Score of 0.132. Performance measures were calculated out of (sub-) sample data, which should improve the robustness of these statistics.

#### Additional model coefficients

```{r}
#| label: fig-odds-other
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "<br><br>"
#| column: page
#| fig-height: 15
#| fig-asp: 1.618
##| out-width: "100%"
# #| fig-width: 12


readRDS(here("data", "251204_dfq_coeff_II.rds")) |>
  filter(!str_detect(term, "year|Interce|iag|acu")) |>
  mutate(grp = str_sub(term, 1, 3)) |> 
  # count(grp)
  # filter(str_detect(term, "diag")) |>
  # filter(str_detect(term, "walk"))
  # arrange(desc(odds)) |>
  # slice_sample(n = 30) |>
  ggplot() +
  geom_hline(yintercept = 1, lty = "dashed", colour = "grey20", alpha = 0.5) +
  geom_pointrange(aes(reorder(term, odds), odds, ymin = lci, ymax = uci, col = grp),  stroke = NA) +
  theme_minimal() +
  coord_flip() +
  scale_y_log10() +
  theme(
     axis.title.x = element_text(size = 9),
    axis.title.y = element_blank(),
    legend.position = "none"
  ) +
  labs(
    y = "\nCase-mix-adjusted odds of admission (log scale)"
  )+
  facet_wrap(vars(grp), scales = "free_y", ncol = 1)


```
