---
title: "Tracking Thresholds for<br>Hospital Admission Via Emergency Department"
subtitle: "A Provider-level analysis<br><br><br>DRAFT REPORT<br>Autumn 2025"
author:
  - name: Steven Wyatt
    email: swyatt@nhs.net
    corresponding: TRUE
    affiliation: The Strategy Unit
  - name: Andrew Jones
    email: swyatt@nhs.net
    corresponding: FALSE
# date: 2024-11-29
format:
  html:
    html-table-processing: none
    # code-fold: true
    # code-tools: true
    embed-resources: true
    grid:
      margin-width: 240px
  pdf: default
reference-location: margin
citation-location: margin
toc: true
toc-title: Contents
toc-location: left
# theme: journal
fig-cap-location: top
number-sections: false
# backgroundcolor: '#FFE0B5'
title-block-banner: '#f8be06'
title-block-banner-color: '#242322'
# mainfont: Gill Sans MT
# mainfont: Helvetica
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

library("gt")
library("here") 
library("arrow")
library("dplyr")
library("purrr")
library("readr") 
library("tidyr")
library("tibble")
library("forcats")
library("ggplot2") 
library("janitor")
library("stringr")
library("gtExtras") 
library("lubridate")

source(here("R", "funs.R"))

# CUT OFF DATE IS MOST RECENT SUNDAY MINUS THREE WEEKS:
param_date_cutoff <- set_cutoff_date(min_buffer_days = 21)

```

# Introduction (placeholder)

This document proposes a method, founded on statistical models, that would allow local areas to monitor whether casemix-adjusted admission thresholds (\*\*for admission via a hospital emergency department) are changing over time.

We have imagined that such information might ultimately be incorporated in a simple software application (app). Using this app, a user might choose any one of 180 Type 1 (24 hour, consultant-led) emergency departments (EDs), and be presented with a range of information to show how admission thresholds are changing over time. The app would include both graphics and text summaries of data-related facts that influence the information they see.

# Methods

## Data source

This work relied on data from the Emergency Care Data Set (ECDS). Each ECDS record represents a patient attendance at an urgent or emergency care department. We selected the subset of records that were coded as unplanned attendances at Type 1 emergency departments (as given by table on NHSE confirmed Type 1 whose provenance is unknown), with an arrival date on or after 1st April 2025.

<!--# We included all streamed patients - since these patients will have faced a triage process which could be included in the admission threshold.  -->

We excluded records where the patient left the ED before treatment was provided; where the patient died during an attendance; and where the individual was found to be dead on arrival at the ED.

Figure X provides an example of how these choices affect the number of emergency attendances included from a given provider (in this case XXXXX) . (from 1st January 2025) !!!!

```{r}
#| label: tbl-where
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "tbd."


readRDS(here("data", "where_matrix.rds")) |> 
  filter(der_provider_site_code == "RWA01") |> 
  mutate(sigma = sum(n)) |>
  rowwise() |> 
  mutate(row_sum = sum(c_across(c(everything(), -c(der_provider_code,der_provider_site_code, n, p, sigma))))) |> 
  ungroup() |> 
  # REMOVE THE ROW CORRESPONDING TO ATTENDANCES FOLLOWING ALL EXCLUSIONS:
  filter(row_sum != 0) |> 
  mutate(n_after_excl = sigma - cumsum(n)) |> 
  # view()
  mutate(n_before_excl = n_after_excl + n) |> 
  select(-c(sigma, p)) |> 
  rename(n_excl = n) |> 
  relocate(c(n_before_excl, n_excl, n_after_excl), .after = row_sum) |> 
  mutate(exclusion = if_else(not_type_1 == 1, "not type 1", "")) |>
  mutate(exclusion = if_else(planned == 1, str_c(exclusion, " planned; "), exclusion)) |>
  mutate(exclusion = if_else(sex_other == 1, str_c(exclusion, " sex-other; "), exclusion)) |>
  mutate(exclusion = if_else(doa == 1, str_c(exclusion, " DOA; "), exclusion)) |>
  mutate(exclusion = if_else(died_ed == 0 & left_died == 1, str_c(exclusion, " left before treatment; "), exclusion)) |>
  mutate(exclusion = if_else(died_ed == 1, str_c(exclusion, " died in ED; "), exclusion)) |>
  select(exclusion, starts_with("n_")) |>
  # view("where")
  gt() |> 
    fmt_auto() |> 
   tab_options(
    data_row.padding = px(1),
    table.font.size = "small"
  ) 

```

## Data quality considerations

<!--# Data quality may be assessed in several dimensions. In this project we came across quality issues that might be characterised as being: 1) accuracy-related; and 2) completeness-related.  -->

Whilst the ECDS is recognised as a high-quality data source, there is yet considerable variation in data quality, both across provider sites and over time. We examine the issues we observed and detail our solutions, below.

### Data quality over time

We chose to use records in which the patient arrived at a ED on or after April 1st 2025, in the belief that this would 1) minimise the variation in data quality that occurs over an extended period; and 2) provide sufficient context in which to interpret current measurements of admission thresholds.

<!--# Whilst this period data quality Unfortunately, we observed a universal step change in the number of missing values in the chief complaint field at the turn of the 2025/26 financial year. This issue influenced some of the choices we made. / limited the choices available to us.  -->

<!--#  lag of a few days? Some 10 providers have incomplete timeseries.... what to do?  -->

<!--#  Providers submit data to the ECDS on a daily basis, and these data may be visible to users of the ECDS a few days later. -->

There was a further balance to achieve when thinking about an end date to our period of study. With an app in mind, we wanted our estimates of admission thresholds to run close to the current date, since the objective would be to present actionable information. However, data quality can be low for very recent data, due to the way data is incrementally flowed from providers to the ECDS. We therefore programmed a rolling end date to our study and a buffer period, such that the most recent data point would cover the last complete week (Mon-Sun) occurring at least 21 days before the computer code was executed\*\*. This buffer period should increase the chances that all sites had submitted high quality data across every field.

I\*\*n the case of the text and graphics presented in this document, the computer code was executed on , which makes the last complete week ...

### Data quality by site

When examining data quality by site, we observed issues that could be characterised as being: 1) accuracy-related; and 2) completeness-related. Here, we will address these dimensions separately.

#### 1. Addressing inaccuracy

Removed from the data-generating processes as we are, we can only infer that data does not represent real-world events. For example, at several sites we observed that almost all patients were coded as having been sent home (FIGURE X). This data is unlikely to reflect the reality at major EDs. In this case, therefore, we left with little choice but to exclude X sites with crude admission rates below XX%.

```{r}
#| label: tbl-accuracy
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "tbd."


# # df_ecds_raw |> colnames()
# #   filter()
# #   count(site_name, ed_discharged)
# df_var_engineering_2of2 |>
#   # select(data) |> 
#   # unnest(data)
#   # slice(-c(3, 14, 20, 22)) |>
#   mutate(p = map_dbl(data, function(df) {
#     df |>
#       count(ed_discharged) |>
#       mutate(p = n/sum(n)) |>
#       filter(ed_discharged == 1) |>
#       pull(p)-1
#   }
#   )) |>
#   mutate(p_adm = abs(p)) |>
#   mutate(site_name = map_chr(data, \(df) df |> distinct(site_name) |> pull() |> as.character())) |>
#   distinct(der_provider_site_code, site_name, p_adm) |>
#   arrange(p_adm) |>
#   # print(n= 11)
#   identity()

```

#### 2. Addressing incompleteness

Approximately x% of the xx million records in our sample (ending XXX September 2025) had missing values in one or more of the fields we had selected for our models. Since our intended logistic regression models would not accept missing values, we had the choice to either: 1) exclude records with missing values; 2) impute missing values; or 3) assign missing values to an explicit missing category (\*\*\*which was possible because the variables with missing values were exclusively categorical).

We opted for the third approach, which allowed us to make use all the records we have. This is especially useful when fitting models for smaller providers, where the number of records (n) that we have may not greatly exceed the number of predictor variables (p) that we wish to include in our regression models. As p –\> n we get closer to an over-fitted, under-powered model, which will be more likely to produce inaccurate predictions and be incapable of distinguishing between geniune effect and random variation.

*The approach also allows the model to make use of the missingness signals. It is important to note that **here we made the assumption that missing data were not missing at random - that there was some systematic reason for the missingness. (((- are they missing at random???))***

If missingness is, however, random, the models might incorporate a false signal, which would compete with other variables for explanatory power, and could ultimately distort the coefficients of interest.

### Addressing non-constant risks

There is another potential hazard which could be amplified by this approach: that of non-constant risks by week. In other words, coding practices at a site - which might include the information that goes *uncoded* - could vary greatly from week to week. And if, for example, admission rates of patients with a missing diagnosis varied greatly from week to week - because of large changes in the casemix that goes uncoded - this would distort the coefficients of interest.

<!--# When thinking about these non-constant risks, we were particularly concerned about the fields indicating the clinician's assessment of the patient (primary diagnosis, chief complaint). These fields will be strongest predictors of admission and ... see below  -->

To highlight results that may have been affected by non-constant risks, we produced a second set of models. The details of these models, and how these models were used to identify possible non-constant risks, can be found in the **Modelling** and **Results** sections respectively).(Testing indicates, however, tht these points will not be a common occurrence.)

## Additonal data processing

Further to the treatment of missing values, we chose to re-engineer *primary diagnosis* and *chief complaint* variables.

<!--#  As these variables indicate the clinician's assessment of a patient's condition, they are likely to be the strongest predictors of admission. -->

Both of these variables, unmodified, consisted of a large number of categories (xx for diagnosis and xx for complaint). This may have caused problems when we came to fit models. (Referring back to Section x, if we can substantially reduce the number of predictors whilst minimising the loss of information, this will be beneficial.)

(Is over fitting a problem here? Since diagnosis - should result in admission or not).

<!--#  We therefore reduced the number of categories in both primary diagnosis and chief complaint fields. -->

We therefore, in each case, grouped together less frequent conditions with the aid of a [3-tier hierarchy (the ECDS "Refset", created by NHSE)](https://digital.nhs.uk/data-and-information/data-collections-and-data-sets/data-sets/emergency-care-data-set-ecds/ecds-guidance). We conducted this grouping at the site level. Groupings may therefore vary, slightly, from site to site.

#### Diagnosis grouping

When dealing with diagnosis, we left the 30 most frequent diagnosis codes un-grouped (thus retaining the most precise diagnosis information). These 30 diagnoses accounted for X% of all (records). We then grouped the remaining diagnoses at a higher (medium-precision) level, and chose the 10 most frequent of these medium-precision diagnosis labels.

Finally, we grouped the remaining diagnoses at the highest, most general level of the hierarchy.

This resulted in an average of XX distinct diagnosis levels. An example for site X is shown in Figure X.

```{r}
#| label: tbl-diagnosis
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "tbd."

# df_var_engineering_2of2 |> 
#   filter(mod_spec == "a") |> 
#   mutate(s = map_dbl(data, function(df){
#                  df |> 
#                    distinct(diag01) |> 
#                    count() |> 
#                    pull()
#   }
#          ))

```

#### Complaint grouping

We followed a similar process with the chief complaint field (note that, unmodified, the detail given in the chief complaint field is equivalent to the second level of the 3-tier hierarchy). We left the most frequent 50 chief complaints un-grouped. (these accounted for X% of chief complaints).

We then grouped the remaining chief complaints at the highest, most general level.

This grouping resulted in an average of XX distinct complaint levels. An example for Site X is shown in Figure Y.

```{r}
#| label: tbl-complaint
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "tbd."

# df_var_engineering_2of2 |> 
#   filter(mod_spec == "b") |> 
#   mutate(s = map_dbl(data, function(df){
#                  df |> 
#                    distinct(complaint) |> 
#                    count() |> 
#                    pull()
#   }
#          ))
  
```

## Variable selection: diagnosis or complaint?

<!--#  When building a statistical model, we aim to select the most relevant and informative variables. In this project invariably have to choose between variables representing similar quantiIn our case, we desired the simplest model thtVariables that duplicate information-->

In terms of variable selection, the major choice concerned two variables that each indicate a clinician's assessment of the patient's condition (which is likely to be the strongest predictor of admission):

-   The *primary diagnosis* variable (hereafter referred to as *diagnosis*) indicates the main condition treated or investigated during the patient's stay in the ED.

-   The *chief complaint* variable (hereafter referred to as *complaint*), on the other hand, refers to "[the primary problem a patient is presenting with, as initially assessed by a clinician](https://www.england.nhs.uk/guidance-for-emergency-departments-initial-assessment/)."

Due to the likely relationship (multicollinearity) between these two variables, we would use only one in a given model. All else being equal, *diagnosis* provides more accurate and precise information about a patient’s condition, and would better allow us to control for casemix. However, if a site has coded *diagnosis* poorly when compared to *complaint*, we questioned if there would be a point at which it would be beneficial to use *complaint* instead of *diagnosis*.

After assessing models using *diagnosis,* and then *complaint,* against the AUROC and Brier Score statistics, we opted to use diagnosis for each site-based model, unless:

-   diagnosis completion rate at a site was below 40% in any one week

    AND

-   the minimum complaint completion rate in any one week was 40% higher than the minimum diagnosis completion rate in any week.

There were x sites that met these criteria. In these cases, we used chief complaint in the model, in place of diagnosis. We therefore had two different model speficiations.

<!--# (It is important to note that the chief complaint and acuity variables are recorded as part of an initial assessement (NHSE, 2022). They are therefore akin to presentation codes.  -->

<!--# should be relevant to the task (of predicting risk of admission) and informative (ultimately allowing the model differentiate between data points)When it came to selecting a variables that indicated a clinician's assessment of a patient's condition, we had a choice to make.  -->

These fields will be strongest predictors of admission and ... see below

One or the other

we have a choice. the field we'd most like to use is diagnosis but we also have chief complaint (a clinician's initial assessment of a patient's condition

in general we use all variables relevant to the task

Our strategy to deal with missing values means that,

<!--#  whilst the model may no longer register missing values, the underlying missingness remains. -->

we may, in some casese, have a large % of explicit missing category. which will not allow the model to

which means most values identical and, as a result, the variables carry little information.

This is fine some information is preferable to no information. But there may be cases where we have a choice.

Indeed, we have a choice to make when deciding

This is the case when thinking about fields indicating the clinician's assessment of the patient (primary diagnosis, chief complaint).

 carry less information and are less likely to differentiate between data points

In key variables we had a choice

fields indicating the clinician's assessment of the patient (primary diagnosis, chief complaint). These fields will be strongest predictors of admission and ... see below

Key variables (we expect to be..

And if the underlying missingness is high An explicit missing may be high % of the values in a variable.

**Choice diagnosis vs complaint**

Site may have a Diagnosis may have a

The strongest predictor is diagnosis – which is our preferred measure of the condition of a patient.

Some sites have high proportion of missing diagnosis values. And complaint better.

the model no longer sees missing values. However, they are s

Whilst our encoding means that the model will no longer register

As far as the model is concerned, there are no longer registers missing values, but this does not change that they are missing we still see these as missingness remains as before

## **Model specification**

### Primary models

Moving from data preparation and variable selection to the modelling stage, we fitted a separate logistic generalised additive model (logistic GAM) for each site.

As noted, we worked with two specifications, identical other than in their use of either the re-engineered *diagnosis* or *complaint* variable, as per section XXX, above.

Otherwise, we included variables to adjust for casemix (as far as possible), namely: *age* (as a spline-based smooth) interacting with *sex*; *acuity*; whether the patient arrived by ambulance or not; and the *referral source*. In addtion, we included time-related variables to control for weekday or weekend attendances, and whether the attendance was during the night (8 pm to 8 am).

We intentionally omitted contextual factors such as the level of crowding and the inpatient occupancy at the time a patient was treated. Instead, we wanted to leave these factors as possible explanations for any changes seen in admission thresholds. If we were to adjust for such effects they would not, in theory, be visible in the time series.

### Models to identify non-constant risks

As noted above (section ???), we produced a second set of models with which we aimed to highlight results from the primary models that may have been affected by non-constant risks.

In preparation for this second set of models, we created an additional binary variable which indicated whether the most influential predictor, primary diagnosis (or chief complaint, where applicable), was missing or not.

An interaction between this new variable and the *week* variable was the only difference between the secondary and primary model specifications. If this interaction exceeded a certain level in any one week, we would flag that week's result as potentially problematic.

## **Reproducibility**

The computer code used in this analysis can be found a [GitHub repository](https://github.com/The-Strategy-Unit/tracking_adm_thresholds) (permissions currently required).

# Results

Show examples

x records, x providers ?? If we're being selective, then why is this necessary?
